PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes

### 2023-02-21 manuscript_00.Rmd
small text changes

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encompassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 35 publications. To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six new cone human region markers. CONCLUSION: The PLAE web application provides the eye community a powerful and quick means to test hypotheses related to gene expression across a highly diverse, community-derived database."

keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

Damage to some of these cell types has been implicated in human eye disease. For example glaucoma is characterized by damage to the retinal ganglion cells while cone and cone-rod dystrophies are characterized by photoreceptor degeneration. Efforts to untangle the gene expression patterns of these diverse cell types with single cell transcriptomics date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003] used RT-PCR approaches to amplify select transcripts in small numbers of cells. Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a]

@@

To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets. We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts. To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019a] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016a] to generate per cell doublet probability scores. To identify cells for removal we used the scores in two ways. First, if an individual cell had a Scrublet score greater than 0.8 (where 1.0 is 100% probability of doublet) and a doubletCell score greater than 1,000,000 (unitless measure where higher value is more likely to be doublet) the cell is labelled as a doublet. Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets. All of these cutoff values were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.

@@

Every study brought into scEiaD was hand curated to identify the unique biological samples, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv2, DropSeq, etc.). We also, where possible, annotated retina region (e.g. macula), sex, and age or developmental stage. Samples with perturbations (genetic or treatment) were not included. To import individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell. We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels and the later are the original cell type labels. We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).

@@

To visualize these top markers we select up to eight genes (ordered by the mean log2 fold change across organisms) per well supported cell type for each and plot the log2 fold change of the gene expression relative to all other cell types (`r fig_cap(name = 'consist_genes_heatmap', display = 'cite')`). We see that human, mouse, and macaque have similar expression patterning of the proposed markers across the well supported cell types. The cell types are arranged by the columns by how related the expression patterns are. We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are near the RPC from which they are derived.

@@

While we have provided data in many forms via the web app at plae.nei.nih.gov, there are many more dissections of the data that can be performed. To facilitate custom testing with the scEiaD database we provide, as an example, a human-specific macula versus peripheral cone differential testing analysis document as supplementary file "pseudobulk_cone_region.freeze01.html". In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region specific manner. This left us with 26678 cells across six studies. Yan et al. previously proposed *Prph2* and *Rs1* as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al. proposed *Calb1* and *Gngt1* as cone region markers in macaque [@pengMolecularClassificationComparative2019a].

>>>>

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    keep_md: true
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high-performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encompassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 35 publications. To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six new cone human region markers. CONCLUSION: The PLAE web application provides the eye community a powerful and quick means to test hypotheses related to gene expression across a highly diverse, community-derived database."

keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'svg') # set output device to svg
```

@@

Damage to some of these cell types has been implicated in human eye disease. For example glaucoma is characterized by damage to the retinal ganglion cells while cone and cone-rod dystrophies are characterized by photoreceptor degeneration. Efforts to untangle the gene expression patterns of these diverse cell types with single cell transcriptomics date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003] used RT-PCR approaches to amplify select transcripts in small numbers of cells. Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a].

@@

To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets [@lunEmptyDropsDistinguishingCells2019a]. We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts. To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019a] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016a] to generate per cell doublet probability scores. To identify cells for removal we used the scores in two ways. First, if an individual cell had a Scrublet score greater than 0.8 (where 1.0 is 100% probability of doublet) and a doubletCell score greater than 1,000,000 (unitless measure where higher value is more likely to be doublet) the cell is labelled as a doublet. Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets. All cutoff values were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.

@@

Every study brought into scEiaD was hand curated to identify the unique biological samples, published paper ID (where available), organ, tissue (e.g. Cornea), source (tissue), single cell platform (e.g. 10Xv2, DropSeq, etc.). We also, where possible, annotated retina region (e.g. macula), sex, and age or developmental stage. Samples with perturbations (genetic or treatment) were not included. To import individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell. We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels and the later are the original cell type labels. We normalized names for CellType, for example changing "MG" to "Muller Glia" and dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).

@@

To visualize these top markers we select up to eight genes (ordered by the mean log2 fold change across organisms) per well supported cell type for each and plot the log2 fold change of the gene expression relative to all other cell types (`r fig_cap(name = 'consist_genes_heatmap', display = 'cite')`). We see that human, mouse, and macaque have similar expression patterning of the proposed markers across the well supported cell types. The cell types are arranged by the columns by how related the expression patterns are. We see how the differentiated cell types of the retina are grouped together apart from the Müller glia, which are near the RPC from which they are derived.

@@

While we have provided data in many forms via the web app at plae.nei.nih.gov, there are many more dissections of the data that can be performed. To facilitate custom testing with the scEiaD database we provide, as an example, a human-specific macula versus peripheral cone differential testing analysis document as supplementary file "pseudobulk_cone_region.freeze01.html". In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region-specific manner. This left us with 26678 cells across six studies. Yan et al. previously proposed *Prph2* and *Rs1* as markers to distinguish macula and peripheral located human cones [@yanCellAtlasHuman2020a] and Peng et al. proposed *Calb1* and *Gngt1* as cone region markers in macaque [@pengMolecularClassificationComparative2019a].
